Reviving stochasticity: uncertainty in SMBH binary eccentricity is unavoidable

### 2023-07-21 papers/paper-eccentricity/paper-eccentricity.tex

\title{Reviving stochasticity: SMBH-binary eccentricity is inherently random}

@@

\author[0000-0001-5721-9335]{Matias mannerkoski}
\affiliation{Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland}

@@

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}

%\drafting{
%    Key results:
%    \begin{enumerate}
%        \item Eccentricity of hard binary depends on deflection angle 
%        \item Mapping $\theta_\mathrm{defl} \rightarrow e_\mathrm{h}$ relatively independent of resolution, however low res explores wider domain space 
%        \item Explored eccentricity space can be full $[0,1]$ range for certain systems 
%        \item Exact values of $e$ from simulations meaningless, need to understand the distribution of 
%        \item Current resolutions appropriate for exploring statistics of eccentricity
%    \end{enumerate}
%    Limited to 250 words.
%}

We study the stochasticity of supermassive black hole (SMBH) binary eccentricity in $N$-body simulations of equal-mass galaxy mergers using the \ketju{} code.
In simulations with realistic, high eccentricity galactic merger orbits, the binary eccentricity is found to be a highly non-linear function of the deflection angle in the SMBH orbit during the final, nearly radial close encounter between the SMBHs before they form a bound binary.
This mapping between the deflection angle and the binary eccentricity is captured using a simple toy model, indicating that it is driven by the interplay between an asymmetric stellar potential and dynamical friction acting on the SMBHs, with no apparent resolution dependence.
Due to the non-linearity of this mapping, in certain merger configurations small, parsec-scale variations in the merger orbit can result in binary eccentricities varying in nearly the full possible range between 0 and 1.
By increasing the number of stellar particles, the reduction in Brownian motion means a smaller range of deflection angles is sampled, thus allowing the binary eccentricity produced by a given merger orbit to converge towards a definite value.
We conclude that the sensitivity of the eccentricity to small-scale perturbations of the merger orbit means that the binary eccentricity is effectively a random quantity when considering realistic galaxy mergers, however we stress that the distribution of these effectively random eccentricities can be constrained using even moderate resolution simulations.

\drafting{
    We present a study of collisionless merger simulations using the recently released \ketju{} code to investigate the stochasticity in the supermassive black hole (SMBH) binary eccentricity at the time a hard binary forms.
    We find a mapping between the hard binary eccentricity and the deflection angle the SMBHs undergo prior to becoming bound that is independent of the ratio between SMBH mass and stellar particle mass. 
    Decreasing the number of stellar particles does however sample a wider domain of deflection angles due to Brownian motion compared to increased stellar particle counts; some merger geometries explore the full $[0,1]$ domain of eccentricity as a result.
    We argue that the SMBH binary eccentricity is not a deterministic function of the initial merger eccentricity, but rather should be understood as a distribution of possible outcomes, and is also most likely present in the observable universe.
    We conclude that mass resolutions currently used to explore the SMBH binary merger process, including their merger timescales, are sufficient to capture the observable population, and that stochasticity in the SMBH binary eccentricity is unavoidable.
}

\drafting{Another version:}

\drafting{
We study the stochasticity of the eccentricity of SMBH binaries formed in $N$-body simulations equal-mass galactic mergers.
In simulations with realistic, high eccentricity galactic merger orbits, the binary eccentricity is found to be a highly non-linear function of the deflection angle in the SMBH orbit during the final, nearly radial close encounter between the SMBHs before they form a bound binary.
This mapping between the deflection angle and binary eccentricity is reproduced with a simple toy model, showing that it is caused by an interplay between the asymmetric stellar potential and dynamical friction acting on the SMBHs.
There is also no apparent resolution dependency in this mapping, with lower particle count simulations simply sampling a wider range of deflection angles due to Brownian motion.
Due to the non-linearity of this mapping, in certain merger configurations small, parsec-scale variations in the merger orbit can result in eccentricities varying in nearly the full possible range between 0 and 1.
Higher particle counts reduce the Brownian motion induced variation in the merger orbit in idealised simulations, allowing the eccentricity produced by a given merger orbit to converge towards a definite value.
However, we argue that the sensitivity of the eccentricity on the small-scale perturbations of the merger orbit means that the binary eccentricity is an effectively random quantity when considering realistic galaxy mergers, but also that the distribution of these effectively random eccentricities can be constrained using even moderate resolution simulations.
}

\drafting{Third version:}

\drafting{
We study the stochasticity of the eccentricity of SMBH binaries formed in $N$-body simulations of equal-mass galactic mergers, run with the \ketju{} code. In simulations with realistic, high eccentricity galactic merger orbits, the binary eccentricity is found to be a highly non-linear function of the final hard scattering, as defined by the deflection angle. The 
mapping between the deflection angle and the binary eccentricity can be reproduced with a simple toy model, showing that it is caused by an interplay between the asymmetric stellar potential and dynamical friction acting on the SMBHs, with no apparent resolution dependency.
At lower particle resolutions the increase in Brownian motion result in a wider sampling of the deflection angles, whereas at higher particle counts a given merger orbit converge towards a more definite eccentricity value.
Due to the non-linearity of this mapping, for certain merger configurations small, parsec-scale variations in the merger orbit can result in final SMBH eccentricities that nearly cover the full $e=[0,1]$ eccentricity range. 
We conclude that the sensitivity of the eccentricity on the small-scale perturbations of the merger orbit means that the binary eccentricity is effectively a random quantity when considering realistic galaxy mergers, however we stress that the distribution of these effectively random eccentricities can be constrained using even moderate resolution simulations.
}

@@

%% Keywords should appear after the \end{abstract} command. 
%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
%% https://astrothesaurus.org
%% You will be asked to selected these concepts during the submission process
%% but this old "keyword" functionality is maintained in case authors want
%% to include these concepts in their preprints.
\keywords{Edit keywords}

@@

%-------------------------- MAIN TEXT STARTS HERE --------------------------%
% INTRODUCTION
\section{Introduction}
Supermassive black holes (SMBHs) are believed to reside at the centres of all massive galaxies \citep[e.g.][]{kormendy2013}.
In the $\Lambda$CDM model as galaxies grow through gas accretion and mergers \citep[e.g.][]{volonteri2003,naab2017} their SMBHs 
are also expected to interact in a three-phase merger process \citep{begelman1980}.

Firstly, dynamical friction \citep{chandrasekhar1943} acts to bring the SMBHs from kiloparsec-scales down to parsec-scale separations, after which the SMBHs form a bound binary with a semimajor axis $a$ and eccentricity $e$ \citep[e.g.][]{milosavljevic2001,merritt2013}. In the second phase, the
SMBH binary separation is reduced through sequential slingshot encounters with the surrounding stellar
distribution \citep{hills1980, hills1983, quinlan1996,rantala2018}. Finally, at small subparsec separations gravitational wave (GW) emission
becomes the dominant mechanism by which the SMBH binary can lose its remaining orbital energy and angular momentum, thus driving the
SMBHs to coalescence \citep[e.g.][]{mannerkoski2019,mannerkoski2022}, where the strength of the GW emission strongly depends on the
binary eccentricity \citep{peters1963,peters1964}.

The complex nature of SMBH coalescence in a galaxy merger setting necessitates the use of numerical techniques \citep[e.g.][]{berentzen2009,khan2011,dosopoulou2017,mannerkoski2019}
to provide quantitative predictions for observational programmes such as ground-based pulsar timing arrays \citep[PTAs, e.g.][]{lentati2015,verbiest2016,arzoumanian2020} and the upcoming Laser Interferometer Space Antenna \citep[LISA, e.g.][]{amaro-seoane2023} space mission.
Understanding how faithfully SMBH binary eccentricity is captured in numerical simulations is thus critical in constraining the SMBH merger rate
that the observational GW community aims to detect.

The SMBH binary merging process and its dependence on eccentricity and resolution has been extensively studied using collisionsless merger simulations \citep[e.g.][]{berentzen2009,vasiliev2015,bortolas2016,gualandris2017,gualandris2022}. 
Recently, \citet{nasim2020} have argued that the scatter in SMBH binary eccentricity observed in gas-free merger simulations is an artefact of poor phase space
discretisation, and that in the real Universe where SMBH masses are far greater than stellar masses $(M_\bullet \gg m_\star)$, SMBH binary eccentricity is a reliably predictable quantity.

In this Letter, we find that the SMBH binary eccentricity is reliably predictable only for particular
initial conditions of the galaxy merger system, and does not in general hold.
Indeed, we find that for realistic galaxy merger orbits where the SMBHs undergo a final near-radial plunge before becoming bound, stochasticity in the final
%initial orbital eccentricity of the system is $e_0>0.90$, stochasticity in the final
SMBH binary eccentricity is physical and unavoidable.

%Previous work by \citet{nasim2020} have argued that scatter in SMBH binary eccentricity observed in gas-free merger simulations is an artefact of poor phase space discretisation, and that in the real Universe where SMBH masses are far greater than stellar masses, $M_\bullet \gg m_\star$, SMBH binary eccentricity is a reliably predictable quantity.

%The first phase, dynamical friction \citep{chandrasekhar1943}, acts to bring the SMBHs from kiloparsec-scale separations to parsec-scale separations, after which the SMBHs form a bound binary system with a semimajor axis $a$ and eccentricity $e$.
%Following the binding of the SMBHs, the SMBH binary separation is reduced through sequential slingshot encounters with the surrounding stellar distribution \citep{hills1980, hills1983, quinlan1996}.
%Chaotic interactions drive the SMBH binary to sub-parsec separation: a regime where gravitational wave (GW) emission becomes the dominant mechanism by which the SMBH binary can lose its remaining orbital energy and angular momentum, resulting in the coalescence of the two SMBHs into a single SMBH.

%It is well known that the efficiency by which GW emission can remove orbital energy and angular momentum from a binary system is highly sensitive to the binary eccentricity \citep{peters1964}, with the binary merger timescale being particularly sensitive.

%The complex nature of SMBH coalescence in a galaxy merger setting necessitates the use of numerical techniques to understand, and to provide quantitative predictions for observational programmes such as pulsar timing arrays \citep[PTAs, e.g.][]{babak2016,falxa2023} and the upcoming Laser Interferometer Space Antenna \citep[LISA, e.g.][]{amaro-seoane2007} mission.
%Understanding how faithfully SMBH binary eccentricity is captured in simulations is thus a critical piece of the puzzle in constraining the SMBH merger rate that the observational GW community aims to detect.

%Previous work by \citet{nasim2020} have argued that scatter in SMBH binary eccentricity observed in gas-free merger simulations is an artefact of poor phase space discretisation, and that in the real Universe where SMBH masses are far greater than stellar masses, $M_\bullet \gg m_\star$, SMBH binary eccentricity is a reliably predictable quantity.

%In this work, we find that SMBH binary eccentricity is reliably predictable only for particular initial conditions of the galaxy merger system, and does not in general hold.
%Indeed, we find that for realistic galaxy merger orbits where the initial orbital eccentricity of the system is $e_0>0.90$, stochasticity in the final SMBH binary eccentricity is present.

@@

\begin{figure*}
    \centering
    \includegraphics{eccentricities.pdf}
    \caption{Eccentricity $e$ (note the non-linear scale) as a function of shifted time $t'=t-t_\mathrm{bound}$ for the $M_\bullet/m_\star=5000$ resolution simulations from the $e_0 = 0.90$ set (blue lines) and $e_0 = 0.99$ set (orange lines). 
    The downwards-pointing arrows indicate the median time when the SMBH binaries becomes hard. 
    In the $e_0=0.90$ set, six simulations have a SMBH binary merger within $50\,\mathrm{Myr}$, shown by the rapid orbit circularisation.}
    \label{fig:eccentricities}
\end{figure*}

We construct a number of idealised galaxy merger simulations, which we evolve with our new version of \ketju{} \citep{mannerkoski2023,rantala2017}, in which the dynamics of
SMBH and stars in a small region around them are integrated with an algorithmically regularised integrator \citep{rantala2020}, whereas the dynamics of the remaining particles
is computed with the \gadget{} \citep{springel2021} fast multiple method (FMM) with second order multipoles. Together with hierarchical time integration this allows for
symmetric interactions and manifest momentum conservation. \ketju{} also includes Post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{blanchet2014}.

Our galaxy models represent the nuclear bulge of a gas-devoid elliptical galaxy, matching the models of \citet{nasim2020}.
Each galaxy consists of a stellar-only Dehnen sphere \citep{dehnen1993} with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the centre of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed.
We test six different mass resolutions 
with a varying number of stellar particles: $N_\star = \{1.0\times10^5,2.5\times10^5, 5.0\times10^5, 1.0\times10^6, 2.0\times10^6, 4.0\times10^6\}$, corresponding to $M_\bullet/m_\star= 1000 \textnormal{--} 40000$. \drafting{See if we include even higher res runs.}

We then construct isolated merger initial conditions by placing two galaxies on two different elliptical orbits with eccentricities of $e_0=0.90$ and $e_0=0.99$, at a fixed
initial separation of $D=3.72\,\mathrm{kpc}$ and semimajor axis $a_0=2.79\,\mathrm{kpc}$.
The radial and tangential velocities of the initial merger setup are consistent with the values reported in \citet{gualandris2022} for the $e_0=0.90$ orbit.
For each orbital configuration, we run ten realisations for each mass resolution, to account for stochasticity caused by the discretised phase space. 
Interactions between stellar particles are softened with a softening length of $\epsilon = 2.5\,\mathrm{pc}$.

@@

The SMBH binary eccentricity for both the $e_0=0.90$ (blue) and $e_0=0.99$ (orange) orbits are shown in \autoref{fig:eccentricities} as a function of shifted time $t' = t - t_\mathrm{bound}$, where $t_\mathrm{bound}$ is the time when the binary orbital energy $E$ becomes permanently negative. 
The median time when each simulation set forms a hard binary is similar between the two sets, as indicated by the arrows in \autoref{fig:eccentricities}.
The $e_0=0.90$ set also demonstrates six SMBH-binary mergers within $50\,\mathrm{Myr}$ of forming a bound binary, seen as a rapid orbit circularisation in \autoref{fig:eccentricities}, which is captured self-consistently using \ketju{}.
The $e_0=0.99$ runs show a wide variation in eccentricity, where $e$ spans almost the entire domain range $[0,1]$.
Even though the $e_0=0.99$ runs have a higher initial merger eccentricity than the $e_0=0.90$ runs, none of the SMBH binaries in the shown $e_0=0.99$ set obtain high enough eccentricities to undergo GW-induced coalescence during the $50\,\mathrm{Myr}$ timespan.

To characterise the scatter in eccentricity, we determine the mean eccentricity $e_\mathrm{h}$ over five orbital periods centred on the orbit within which the SMBH binary has become hard.
The binary hardening radius $a_\mathrm{h}$ is defined as \citep[e.g.][]{merritt2006}
\begin{equation}\label{eq:ahard}
    a_\mathrm{h} = \frac{q}{(1+q)^2} \frac{r_\mathrm{m}}{4} = \frac{r_\mathrm{m}}{16},
\end{equation}
where $r_\mathrm{m}=r(m<2M_{\bullet,1})$ is the influence radius and $q$ is the mass ratio between the SMBHs (for our simulations $q=1.0$). 
We characterise the inter-simulation eccentricity scatter in the mean values of $e_\mathrm{h}$ with the standard deviation, denoted as $\sigma_e$.

We show the dependence of the inter-simulation eccentricity standard deviation on mass resolution for the $e_0=0.90$ orbit in \autoref{fig:convergence} with blue circle markers, and for the $e_0=0.99$ orbit using orange square markers.

For the $e_0=0.90$ orbit, the convergence of $\sigma_e$ scales as $1/\sqrt{N_{\star,\mathrm{tot}}}$, where $N_{\star,\mathrm{tot}}$ is the total number of stellar particles in the merger, in agreement with the results of \citet{nasim2020}. 
The variation in eccentricity does not show the same scaling for the $e_0=0.99$ orbit as the $e_0=0.90$ orbit. 
For mass resolutions $M_\bullet/m_\star<10000$, the value of $\sigma_e$ is almost constant, before significantly dropping at higher mass resolutions. \drafting{Add about 4M runs when done}

@@

To understand the observed scatter in eccentricity, we investigate the binary binding process in each simulation. 
Whilst the two SMBHs are unbound, they undergo a number of hyperbolic orbits influenced by the galactic merger potential.
At each pericentre passage, the two SMBHs scatter off each other with some deflection angle $\theta_\mathrm{defl}$, defined:
\begin{equation}\label{eq:theta}
    \theta_\mathrm{defl} = 2 \arctan \left(\frac{GM}{L\sqrt{2E}} \right)
\end{equation}
where $M=M_{\bullet,1}+M_{\bullet,2}$, and $L$ is the magnitude of the SMBH 
binary angular momentum vector and $E$ the binary orbital energy at the time of the pericentre passage \citep{binney2008}.
The deflection angle is a function of the impact parameter $b$ of the SMBH trajectory. However as $\theta_\mathrm{defl}$ can be more robustly measured from the simulation data, we instead
opt using it for our analysis.

Each pericentre passage is associated with a deflection angle, however we find that it is the first extreme deflection angle, associated with a hard scattering event, that is important.
To quantify an extreme deflection angle, we set as $\theta_\mathrm{defl}$ the first angle that exceeds $\theta_{\mathrm{defl,min}}=30\degr$. \drafting{Is this detail necessary?}

In general, this corresponds to the second or third actual pericentre passage of the SMBHs in the simulations.
An example of $\theta_\mathrm{defl}$ is shown for two realisations of the $e_0=0.90$ ($e_0=0.99$) orbit in the left (right) panel of \autoref{fig:orbit}, which occurs between the arrow heads along the SMBH trajectory in each inset panel. 
We observe evidence for a relationship between the deflection angle $\theta_\mathrm{defl}$ and the resulting eccentricity at the time the SMBH binary becomes hard, as shown in \autoref{fig:theta_e}.

\begin{figure*}
    \centering
    \includegraphics{orbit}
    \caption{Orbits from $t=0\,\mathrm{Myr}$ to a time shortly after $t_\mathrm{bound}$ of a single SMBH in two representative realisations of the $e_0=0.90$ mergers (left) and the $e_0=0.99$ mergers (right), which by symmetry of the equal mass system is a reflection of the second BH orbit about the $x=-z$ line. 
    Each line is coloured according to the shifted time $t'=t-t_\mathrm{bound}$ of the simulation, however the colours for $|t'|>2\,\mathrm{Myr}$ is constant for visual clarity. % of the deflection-phase of the evolution. 
    In each inset panel, $\theta_\mathrm{defl}$ is the deflection at the pericentre between the arrows.
    The circle marker indicates the SMBH position just before a bound binary is formed.}
    \label{fig:orbit}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics{theta_e_sim_and_model}
    \caption{
    Scattering deflection angle $\theta_\mathrm{defl}$ and the resulting binary eccentricity
    in the $e_0=0.90$ (left) and $e_0=0.99$ (right) simulations,
    compared to fitted model curves.
    The solid lines show the results from the toy model with $e_\mathrm{s}=0.91$,
    while the semitransparent lines show the results for  $e_\mathrm{s}=0.90$ and $e_\mathrm{s}=0.92$.
    The $e_0=0.90$ model curves have been shifted left by $14\degr$
    to better match the simulation data, with the shift also indicated by an arrow.
    The marginal histograms show kernel density estimates of the simulation data.
    The inset panel for the $e_0=0.90$ data shows a zoom-in with $\theta_\mathrm{defl}=[52\degr, 73\degr]$ and $e_\mathrm{h}=[0.96, 1.00]$ for clarity.
    }
    \label{fig:theta_e}
\end{figure*}

\subsection{Reproducing the eccentricity behaviour with a toy model}\label{ssec:toy}

>>>>

\title{Reviving stochasticity: uncertainty in SMBH binary eccentricity is unavoidable}

@@

\author[0000-0001-5721-9335]{Matias Mannerkoski}
\affiliation{Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland}

@@

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
We study supermassive black hole (SMBH) binary eccentricity of equal-mass galaxy mergers in $N$-body simulations with the \ketju{} code, which combines the \gadget{} fast multipole gravity solver with accurate regularized integration and post-Newtonian corrections around SMBHs.
In simulations with realistic, high eccentricity galactic merger orbits, the binary eccentricity is found to be a non-linear function of the deflection angle in the SMBH orbit during the final, nearly radial close encounter between the SMBHs before they form a bound binary.
This mapping between the deflection angle and the binary eccentricity has no apparent resolution dependence in our simulations spanning the resolution range of $1\times10^5$--$8\times10^6$ particles per galaxy.
The mapping is also captured using a simple model with an analytic potential, indicating that it is driven by the interplay between a smooth asymmetric stellar background potential and dynamical friction acting on the SMBHs.
Due to the non-linearity of this mapping, in certain merger configurations small, parsec-scale variations in the merger orbit can result in binary eccentricities varying in nearly the full possible range between $e=0$ and $e=1$.
In idealized simulations, such variations are caused by finite resolution effects,
and convergence of the binary eccentricity can be achieved with increasing resolution.
However, in real galaxies, other mechanisms such as nuclear gas and substructure that perturb the merger orbit are likely to be significant enough for the binary eccentricity to be effectively random.
Our results indicate that the distribution of these effectively random eccentricities can be studied using even moderate resolution simulations.

@@

%% Keywords should appear after the \end{abstract} command. 
%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
%% https://astrothesaurus.org
%% You will be asked to selected these concepts during the submission process
%% but this old "keyword" functionality is maintained in case authors want
%% to include these concepts in their preprints.
%\keywords{Galaxy mergers (608); Supermassive black holes (1663); Eccentricity (441); N-body simulations (1083); }

@@

%-------------------------- MAIN TEXT STARTS HERE --------------------------%
% INTRODUCTION
\section{Introduction}
Supermassive black holes (SMBHs) are believed to reside at the centers of all massive galaxies \citep[e.g.][]{kormendy2013}.
In the $\Lambda$CDM model as galaxies grow through gas accretion and mergers \citep[e.g.][]{volonteri2003,naab2017} their SMBHs 
are also expected to interact in a three-phase merger process \citep{begelman1980}.

Firstly, dynamical friction \citep[DF,][]{chandrasekhar1943} acts to bring the SMBHs from kiloparsec-scales down to parsec-scale separations, after which the SMBHs form a bound binary with a semimajor axis $a$ and eccentricity $e$ \citep[e.g.][]{milosavljevic2001,merritt2013}. In the second phase, the
SMBH binary separation is reduced through sequential slingshot encounters with the surrounding stellar
distribution \citep{hills1980, hills1983, quinlan1996,rantala2018}. Finally, at small subparsec separations gravitational wave (GW) emission
becomes the dominant mechanism by which the SMBH binary can lose its remaining orbital energy and angular momentum, thus driving the
SMBHs to coalescence, where the strength of the GW emission strongly depends on the
binary eccentricity \citep{peters1963,peters1964,mannerkoski2019}.

The complex nature of SMBH coalescence in a galaxy merger setting necessitates the use of numerical techniques \citep[e.g.][]{berentzen2009,khan2011,dosopoulou2017,mannerkoski2019}
to provide quantitative predictions for ongoing observational programs such as ground-based pulsar timing arrays \citep[PTAs,][]{agazie2023,2023Antoniadis,2023Xu,zic2023}
and the upcoming Laser Interferometer Space Antenna \citep[LISA, e.g.][]{amaro-seoane2023}.
Understanding how faithfully SMBH binary eccentricity is captured in numerical simulations is thus critical in constraining the SMBH merger rate
that the observational GW community aims to detect.

The SMBH binary merging process and its dependence on eccentricity and resolution has been extensively studied using collisionsless merger simulations \citep[e.g.][]{berentzen2009,vasiliev2015,bortolas2016,gualandris2017,gualandris2022}. 
Recently, \citet{nasim2020} have argued that the substantial scatter in SMBH binary eccentricity observed in gas-free merger simulations is an artefact of poor phase space
discretization, and that in the real Universe where SMBH masses are far greater than stellar masses $(M_\bullet \gg m_\star)$, SMBH binary eccentricity is a reliably predictable quantity.

In this Letter, we find that for realistic galaxy merger orbits the scatter in the final SMBH binary eccentricity is due to the 
physical sensitivity of the final eccentricity to small perturbations of the final  nearly radial plunging trajectory of the SMBHs before they become bound, even in the infinite resolution limit.
As a result, we argue that the stochasticity of the binary eccentricity is an unavoidable physical feature of realistic galaxy mergers, at least in the equal-mass case.

\begin{figure*}
    \centering
    \includegraphics{eccentricities.pdf}
    \caption{Binary eccentricity $e$ (note the non-linear scale) and shifted time $t'=t-t_\mathrm{bound}$ for the $M_\bullet/m_\star=5000$ resolution simulations from the $e_0 = 0.90$ set (blue lines) and $e_0 = 0.99$ set (orange lines). 
    The downwards-pointing arrows indicate the median time when the SMBH binaries becomes hard. 
    In the $e_0=0.90$ set, six simulations have a binary merger within $50\,\mathrm{Myr}$.}
    \label{fig:eccentricities}
\end{figure*}

@@

We construct a number of idealized galaxy merger simulations, which we evolve with our new version of the \ketju{} code \citep{mannerkoski2023,rantala2017}.
The dynamics of
SMBHs, and stars in a small region around them, are integrated with an algorithmically regularized integrator \citep{rantala2020}, whereas the dynamics of the remaining particles
is computed with the \gadget{} \citep{springel2021} fast multiple method (FMM) with second order multipoles. Together with hierarchical time integration this allows for
symmetric interactions and manifest momentum conservation. \ketju{} also includes post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{blanchet2014}.

Our galaxy models represent the nuclear bulge of a gas-devoid elliptical galaxy, matching the models of \citet{nasim2020}.
Each galaxy consists of a stellar-only \citet{dehnen1993} sphere with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the center of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed.
We test seven different mass resolutions 
with a varying number of stellar particles: $N_\star = \{1.0\times10^5,2.5\times10^5, 5.0\times10^5, 1.0\times10^6, 2.0\times10^6, 4.0\times10^6, 8.0\times10^6\}$, corresponding to $M_\bullet/m_\star= 1000 \textnormal{--} 80000$.

We then construct isolated merger initial conditions by placing two galaxies on two different elliptical orbits with eccentricities of $e_0=0.90$ and $e_0=0.99$, at a fixed
initial separation of $D=3.72\,\mathrm{kpc}$ and a semimajor axis of $a_0=2.79\,\mathrm{kpc}$.
The $e_0=0.90$ merger orbit matches that used by \citet{nasim2020}, with the radial and tangential velocities being consistent with the values reported in \citet{gualandris2022}.
For each orbital configuration, we run ten realizations for each mass resolution, to account for stochasticity caused by the discretized phase space. 
Interactions between stellar particles are softened with a softening length of $\varepsilon = 2.5\,\mathrm{pc}$, and the \ketju{} region radius is $r_{\rm ketju}=3\varepsilon = 7.5\,\mathrm{pc}$.

@@

The SMBH binary eccentricity for both the $e_0=0.90$ (blue) and $e_0=0.99$ (orange) orbits are shown in \autoref{fig:eccentricities} as a function of shifted time $t' = t - t_\mathrm{bound}$, where $t_\mathrm{bound}$ is the time when the binary orbital energy $E$ becomes permanently negative. 
The median time when each simulation set forms a hard binary is similar between the two sets, as indicated by the arrows in \autoref{fig:eccentricities}.
The $e_0=0.90$ set also demonstrates six SMBH-binary mergers within $50\,\mathrm{Myr}$ of forming a bound binary, seen as a rapid orbit circularization in \autoref{fig:eccentricities}, which is captured self-consistently using \ketju{}.
The $e_0=0.99$ runs show a wide variation in eccentricity, where $e$ spans almost the entire domain range $e=[0,1]$.
Even though the $e_0=0.99$ runs have a higher initial merger eccentricity than the $e_0=0.90$ runs, none of the SMBH binaries in the shown $e_0=0.99$ set obtain high enough eccentricities to undergo GW-induced coalescence during the $50\,\mathrm{Myr}$ timespan.

To characterize the scatter in eccentricity, we determine the mean eccentricity $e_\mathrm{h}$ over five orbital periods centered on the orbit within which the SMBH binary has become hard.
The binary hardening radius $a_\mathrm{h}$ is defined as \citep[e.g.][]{merritt2006}:
\begin{equation}\label{eq:ahard}
    a_\mathrm{h} = \frac{q}{(1+q)^2} \frac{r_\mathrm{m}}{4} = \frac{r_\mathrm{m}}{16},
\end{equation}
where $r_\mathrm{m}=r(m<2M_{\bullet,1})$ is the influence radius and $q$ is the mass ratio between the SMBHs (for our simulations $q=1.0$). 
We characterize the inter-simulation eccentricity scatter in the mean values of $e_\mathrm{h}$ with the standard deviation, denoted as $\sigma_e$.

We show the dependence of $\sigma_e$ on mass resolution for the $e_0=0.90$ orbit in \autoref{fig:convergence} with blue circle markers, and for the $e_0=0.99$ orbit using orange square markers.

For the $e_0=0.90$ orbit, the convergence of $\sigma_e$ scales as $1/\sqrt{N_{\star,\mathrm{tot}}}$, where $N_{\star,\mathrm{tot}}$ is the total number of stellar particles in the merger.
The values of $\sigma_e$ we report are quantitatively similar to the values found by \citet{nasim2020} for the same system. 
The variation in eccentricity does not show the same scaling for the $e_0=0.99$ orbit as the $e_0=0.90$ orbit. 
For mass resolutions $M_\bullet/m_\star \leq 10000$, the value of $\sigma_e$ is almost constant, before significantly dropping at higher mass resolutions.

\begin{figure*}
    \centering
    \includegraphics{orbit}
    \caption{Orbits from $t=0\,\mathrm{Myr}$ to a time shortly after $t_\mathrm{bound}$ of a single SMBH in two representative realizations of the $e_0=0.90$ mergers (left) and the $e_0=0.99$ mergers (right), which by symmetry of the equal mass system is a reflection of the second BH orbit about $x=-z$. 
    The line gradient shows the shifted time $t'=t-t_\mathrm{bound}$, however the coloring for $|t'|>2\,\mathrm{Myr}$ is constant for visual clarity. 
    In each inset panel, $\theta_\mathrm{defl}$ is the deflection angle at the pericenter between the arrows.
    The circle marker indicates the SMBH position when a bound binary is formed.}
    \label{fig:orbit}
\end{figure*}

@@

To understand the observed scatter in eccentricity, we investigate the binary binding process in each simulation. 
Before the SMBHs are bound, the primary influence on the motion is from the galactic merger potential; two realizations for both merger orbits are shown in \autoref{fig:orbit}.
The initial stages of the orbit show indiscernible variation between realizations, however after a particularly strong interaction between the SMBHs during a pericenter passage, the orbits deviate between realizations (inset panels, \autoref{fig:orbit}), and evolve to different binary eccentricities.

To quantify the strength of the hard-scattering process that randomizes the SMBH orbit, we measure the effective two-body deflection angle, defined as:
\begin{equation}\label{eq:theta}
    \theta_\mathrm{defl} = 2 \arctan \left(\frac{GM}{L\sqrt{2E}} \right) = 2\arctan\left( \frac{b_{90}}{b} \right)
\end{equation}
where $M=M_{\bullet,1}+M_{\bullet,2}$, $L$ is the magnitude of the SMBH 
system angular momentum vector, $E$ the orbital energy at the time of the pericenter passage, $b$ the effective impact parameter, and $b_{90}$ the $90\degr$ deflection radius \citep{binney2008}.

We observe clear evidence for a relationship between the deflection angle $\theta_\mathrm{defl}$ and the resulting eccentricity at the time the SMBH binary becomes hard, as shown with the data points in \autoref{fig:theta_e}.

\begin{figure*}
    \centering
    \includegraphics{theta_e_sim_and_model}
    \caption{
    Deflection angle $\theta_\mathrm{defl}$ and resulting binary eccentricity
    in the $e_0=0.90$ (left) and $e_0=0.99$ (right) simulations,
    compared to fitted model curves.
    The solid lines show the results from the analytic model with $e_\mathrm{s}=0.91$,
    while the semitransparent lines show the results for  $e_\mathrm{s}=0.90$ and $e_\mathrm{s}=0.92$.
    The $e_0=0.90$ model curves have been shifted left by $14\degr$
    to better match the simulation data, with the shift indicated by an arrow.
    The marginal histograms show kernel density estimates of the simulation data.
    The inset panel for the $e_0=0.90$ data shows a zoom-in with $\theta_\mathrm{defl}=[52\degr, 73\degr]$ and $e_\mathrm{h}=[0.96, 1.00]$.
    }
    \label{fig:theta_e}
\end{figure*}

\subsection{Reproducing the eccentricity behavior with a simple model}\label{ssec:toy}

The dependency of the binary eccentricity on the 
deflection angle $\theta_\mathrm{defl}$ during the binary formation
can be reproduced using a simple analytic model,
which includes only the essential components relevant to this process.
Taking the orbit to lie in the $xy$-plane,
the relative motion of the two equal-mass SMBHs in this model follows the 
equation of motion
\begin{equation} \label{eq:toy_model_eom}
\ddot{\vb{x}} = -\frac{2 G M_\bullet}{|\vb{x}|^3} \vb{x} + \vb{a}_\mathrm{bg} + \vb{a}_\mathrm{DF},
\end{equation}
where $\vb{x}=(x,y)$ is the separation vector of the SMBHs,
$M_\bullet = 10^8\,\Msun$ is the mass of a single SMBH,
$\vb{a}_\mathrm{bg}$ the acceleration due to the asymmetric background potential,
and $\vb{a}_\mathrm{DF}$ is the acceleration due to DF.
